---
title: Introduction to R programming
latest-revision: October 3rd, 2023
author: Vittorio Zampinetti
time: 3h
output: pdf_document
---

# Introduction

## R and RStudio

- [**R**](https://www.r-project.org/): programming language, tailored for
statisticians
- [**RStudio**](https://www.rstudio.com/): development environment, software for
editing and running R (and Python) applications
  
  It's going to change name soon, see [Posit](https://posit.co/)
  
### Installation

First install R programming language (from [here](https://cran.rstudio.com/)),
then RStudio environment (from
[here](https://www.rstudio.com/products/rstudio/download/#download)).

## R syntax

Here a brief overview of the main components of the R language.

*Note: this is not a complete guide on the language nor on programming in*
*general, but rather a quick-start introduction with the most used operations*
*and structures, which assumes some very basic programming skills.*
*For all the rest, Google is your friend (it really is).*

### Basic operations

R can be used as a fully featured calculator, typically directly
from the console (see bottom panel in RStudio).

Some examples:
```{r}
2 + 2 # calculator-style
c(1, 2, 3) * 4 # vector operations
t(matrix(c(c(1, 2, 3, 4)),
         nrow = 2, ncol = 2)) # matrix operations (transpose)
```
As any other programming language, it allows for variables declaration.
Variables hold a value which can be assigned, modified, and used for
other computations.
```{r}
a <- 2 # assign values to variables
b = a + 2 # equal sign also valid, but...
b
```
The arrow sign `<-` is the traditional sign for the assignment operation, 
but also `=` works. Check [this examples](#arrow-sign) to see why `<-` 
is recommended.

Conditional statements and loops have straightforward syntax (similar to C/C++
, but more compact).
```{r}
if (b == 4 && a > 0) { # if-else statement
  out <- ""
  for (i in 1:b) { # for loop
    out <- paste(out, class(b)) # concatenate two or more strings
  }
  print(out) # print string to output
} else {
  stop("error") # stop program and exit
}
```

While `<-` is only used for variable assignment, i.e. filling up the variable
allocated space with some value, the `=` is both used for assignment and
function argument passing (as introduced in [Functions](#functions)).

## Functions

Functions are called with the `()` operator and present named arguments.
```{r}
# get Normal samples with 0 (default) mean and 10 standard deviation
array <- rnorm(10, sd = 10) # sd is an argument,
                            # the mean is left to the default 0
array
```
You can explicitly call the arguments to which you want to pass the parameters
(e.g. sd of the Gaussian). In this case it is necessary, because the second
position argument is the mean, but we want it to be the default.

### Definition

To define a custom function

```{r}
# implements power operation
#'
#' @param x a real number
#' @param n a natural number, defaults to 2
#' @return n-th power of x
pow <- function(x, n = 2) {
  ans <- 1
  if (n > 0) {
    for (i in 1:n) {
      ans <- ans * x
    }
  } else if (n == 0) {
    # do nothing, ans is already set to 1
  } else {
    stop("error: n must be non-negative")
  }
  return(ans)
}

print(paste("3^5 = ", pow(3, 5),
            ", 4^2 = ", pow(4),
            ", 5^0 = ", pow(5, 0)))
```
The `return()` call can be omitted (but it's better not to).

```{r, eval=FALSE}
pow <- function(a, b) {
  ans <- 1
  # ...
  # last expression is returned
  ans # this is equivalent to write return(ans)
}
```


## Data types

Every variable in R is an object, which holds a value (or collection of values) and some other attributes, properties or methods.

### Base types

There are 5 basic data types:

- **numeric** (real numbers)
- **integer**
- **logical** (aka boolean)
- **character**
- **complex** (complex numbers)

#### Numeric
When you write numbers in R, they are going to
default to numeric type values

```{r}
a <- 3.4 # decimal
b <- -.30 # signed decimal
c <- 42 # also without dot, it's numeric

print(paste0("data types | a:", class(a),
            ", b:", class(b),
            ", c:", class(c)))

```
#### Integer

Integer numbers can be enforced typing an `L` next to the digits.
Casting is implicit when the result of an operation involving integers
is not an integer

```{r}
int_num <- 50L # 50 stored as integer

non_casted_num <- int_num - 2L # result is exactly 48, still int
casted_num <- int_num / 7L # implicit cast to numeric type to store decimals
print(paste(class(int_num), class(non_casted_num),
            class(casted_num), sep = ", "))
```

#### Logical

The `logical` type can only hold `TRUE` or `FALSE` (in capital letters)
```{r}
bool_a <- FALSE
bool_b <- T # T and F are short for TRUE and FALSE
bool_a | !bool_b # logical or between F and not-T
```
You can test the value of a boolean also using `0,1`
```{r}
bool_a == 0 # if 'A' is not equal to 'not B', raise an error
```
and a sum between logical values, treats `FALSE = 0` and `TRUE = 1`,
which is useful when counting true values in logical arrays
```{r}
bool_a + bool_b + bool_b # FALSE + TRUE + TRUE (it's not an OR operation)
```
#### Character

A character is any number of characters enclosed in quotes `' '` or double
quotes `" "`.
```{r}
char_a <- "," # single character
char_b <- "bird" # string
char_c <- 'word' # single quotes
full_char <- paste(char_b, char_a, char_c) # concatenate chars
class(full_char) # still a character
```

#### Complex

```{r}
complex_num <- 5 + 4i
Mod(complex_num) # try all complex operations, e.g. modulus
```

#### Special values

- `NA`: "not available", missing value
- `Inf`: infinity
- `NaN`: "not-a-number", undefined value

```{r}
missing_val <- NA
is.na(missing_val) # test if value is missing
```

Every operation involving missing values, will output `NA`
```{r}
missing_val == NA # cannot use ==
```
```{r}
print(paste(
  "1/0 = ", 1 / 0,
  ", 0/0 = ", 0 / 0
))
```
These special values are not necessarily unwanted, but they require
extra care. E.g. `Inf` can appear also in case of numerical *overflow*.
```{r}
exp(1000)
```

#### Conversion

Variables types can be converted with `as.<typename>()`-like functions, as long
as conversion makes sense. Some examples:

```{r}
v <- TRUE
w <- "0"
x <- 3.2
y <- 2L
z <- "F"
cat(paste(
  paste(x, as.integer(x), sep = " => "), # from numeric to integer
  paste(y, as.numeric(y), sep = " => "), # from integer to numeric
  paste(y, as.character(y), sep = " => "),  # from integer to character
  paste(w, as.numeric(w), sep = " => "),  # from number-char to numeric
  paste(v, as.numeric(v), sep = " => "), # from logical to numeric
  sep = "\n"
))
```

```{r, error=TRUE}
as.numeric(z)  # from character to numeric (coercion warning - NA)
```

### Vectors and matrices

#### Vectors

Vectors are build with the `c()` function. 
A vector holds values of the same type.

```{r}
vec1 <- c(4, 3, 9, 5, 8)
vec1
```
Vector operations and aggregation of values is as intuitive as it can be.
```{r}
vec2 <- vec1 - 1 # subtract 1 to all values (broadcast)
sum(vec1) # sum all values in vec1
mean(vec2) # compute the mean
sort(vec1, decreasing = TRUE) # sort elements in decreasing order
```

Conversion is still possible and it's made element by element.
```{r}
char_vec <- as.character(vec1) # convert every value in vec1 to char
char_vec
```

Range vectors (unit-stepped intervals) are built with `start:end` syntax.
Note: the type of range vectors is `integer`, not `numeric`.
```{r}
x_range <- 1:10
class(x_range)
```

They are particularly useful in loops statements:
```{r}
vec3 <- c() # declare an empty vector
# iterate all the indices along vec1
for (i in 1:length(vec1)) {
  vec3[i] <- vec1[i] * i # access with [idx]
}
vec3
```

Vector elements are selected with square
brackets `[]`. Putting vectors inside brackets performs
slicing

```{r}
vec1[1:3] # first 3 elements
vec1[c(1, 3)] # only first and third element
vec1[-c(1:3)] # all but elements 1 to 3
vec1[seq(1, length(vec1), 2)] # odd position elements
```

To find an element in a vector and get its index/indices,
the `which()` function can be used

```{r}
which(vec1 == 3)
which(vec1 < 5)
```

And finally, to filter only values that satisfy a 
certain condition, we can combine `which` with 
splicing.

```{r}
vec1[which(vec1 >= 5)]

# or, equivalently, using logical masking
vec1[vec1 >= 5]
```

#### Matrices

Matrices are built with `matrix()`
```{r}
mat1 <- matrix(1:24,
               nrow = 6, ncol = 4)
mat1 # filled column-wise (default)
```

```{r}
mat2 <- matrix(1:24,
               nrow = 6, ncol = 4, byrow = TRUE)
mat2 # filled row-wise
```

```{r}
dim(mat2) # get dimensions
c(nrow(mat2), ncol(mat2)) # get number of rows and cols separately

# or, equivalently
dim(mat2)[1] # nrow
```
All indexing operations available on vectors, are also available on 
matrices

```{r}
mat2[1, 1] # element 1,1
mat2[3, ] # third row (empty space for all elements)
mat2[1:2, 1:2] # upper left 2x2 sub-matrix
t(mat2) # transposed matrix
```

Operations with matrix and vectors can be both element-wise
and matrix operations (e.g. scalar product).
Note that a vector built with `c()` is a column vector by default.
Some examples:

```{r}
diagonal_mat <- diag(nrow = 4) # 4x4 identity matrix
 # element by element
diagonal_mat * 1:2 # note: 1:2 is repeated to match the matrix dimensions
diagonal_mat %*% seq(2, 8, 2) # matrix multiplication (4,4) x (4, 1) -> (4, 1)

v1 <- 1:4
v2 <- 4:1
v1 %*% v2 # here v1 is implicitly converted to row vector
```

#### Arrays 

Arrays are multi-dimensional vectors (generalization
of a matrix with more than two dimensions).
They work pretty much like matrices.

```{r}
arr1 <- array(1:24, dim = c(2, 4, 3))
arr1
arr1[2, 1, 3] # get one element
sliced_arr <- arr1[, 2, ] # slice at column 2
sliced_arr
dim(sliced_arr) # reduces ndims by one (dimension selected is dropped)
```

### Lists and dataframes

Lists are containers that can hold different data types. Each entry,
which can even be another list, has a position in the list and can also
be named.

```{r}
list1  <- list(1:3, TRUE, x = c("a", "b", "c"))
list1
list1[[3]] # access with through index
list1$x # access through name
```

Dataframes are collections of columns that have the same length.
Contrarily to matrices, columns in dataframes can be of different types.
They are the most common way of representing structured data and 
most of the dataset will be stored in dataframes.

```{r}
df1 <- data.frame(x = 1, y = 1:10,
           char = sample(c("a", "b"), 10, replace = TRUE))

df1 # x was set to just one value and gets repeated ('recycled')
df1[[2]] # access through column index
df1$x # access through column name
df1[, 3] # access with matrix-style index
df1[2:4, ] # can also select subset of rows
```

The `dplyr` library provides another dataframe object (called *tibble*) which has all 
the effective features of Base R `data.frame` and none of the 
deprecated functionalities. It's simply a newer version of dataframes (therefore
recommended over the old one).

```{r}
library("tibble")
tibble(x = 1:15, y = 1, z = x / y) # tibble dataframe
```

For more information on tibble and its advantages with respect to traditional
dataframes, type `vignette("tibble")` in an R console.
Notice that you can convert datasets to tibble with `as_tibble()`, while 
with `as.data.frame()` you will get a Base R dataframe.

## Data manipulation

Now that we know what a dataframe is and how it is generated,
we can focus on data manipulation.

The `dplyr` library provides an intuitive way of working with datasets.
For instance, let's consider the `mtcars` dataset.

```{r}
library(dplyr)

mtcars$modelname <- rownames(mtcars) # name column with models
mtcars <- as_tibble(mtcars) # convert to tibble
mtcars # display the raw data
```
Let's say we want to get the cars with more than 100 hp, and we are just
interested in the car model name and we want the data to be sorted in alphabetic
order.

```{r}
mtcars %>% # send the data into the transformation pipe
  dplyr::filter(hp > 100) %>% # filter rows with hp > 100
  dplyr::select(modelname) %>% # filter columns (select only modelname col)
  dplyr::arrange(modelname) # display in alphabetic order
```
There are many other dplyr functions for data transformation. This useful
[cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/data-transformation.pdf)
summarizes most of them for quick access.

## Plotting

In base R plots are built with several calls to functions, each of which
edit the current canvas. For instance, to plot some points and a line:

```{r}
# generate synthetic data
n_points <- 20
x <- 1:n_points
y <- 3 * x + 2 * rnorm(n_points)

plot(x, y)
abline(a = 0, b = 3, col = "red")
```

However, the `ggplot` library is now the new standard plotting library.
In ggplot, a plot is decomposed in three main components: data, coordinate 
system and visual marks, called *geoms*.
The plot is built by stacking up layers of visualization objects.
Data is in form of dataframes and the columns are selected in the aesthetics
arguments.

The same plot shown before can be drawn with ggplot in the following way.

```{r}
library(ggplot2)
gg_df <- tibble(x = x, y = y)

ggplot(gg_df) +
  geom_point(mapping = aes(x, y)) +
  geom_abline(mapping = aes(intercept = 0, slope = 3), color = "red")
```

This is just a brief example. More will be seen in the next lessons.
Check out this
[cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/data-visualization.pdf)
for quick look-up on ggplot functions.

## Examples: plot and data manipulation

Combining altogether, here a data visualization workflow on the 
[Gapminder](https://www.gapminder.org/data/) dataset.

```{r}
library(gapminder)
# have a quick look at the Gapminder dataset
str(gapminder)
```
A factor, which we haven't seen yet, is just a data-type characterizing a
discrete categorical variable; the levels of a factor describe how many
distinct categories it can take value from (e.g. the variable `continent` takes
values from the set `{Africa, Americas, Asia, Europe, Oceania}`).

Let's say we want to compare the GDP per capita of some
different countries (Italy, Japan, Brasil and Ethiopia), 
plotted against time (year by year).

```{r}
# transform the dataset according to what is necessary
wanted_countries <- c("Italy", "Japan", "Brazil", "Ethiopia")
gapminder %>%
  dplyr::filter(country %in% wanted_countries) %>%
  # now feed the filtered data to ggplot (using the pipe op)
  ggplot() +
    geom_line(aes(year, gdpPercap, color = country))
```

If we want to add some information about the same measure over the whole
continent, showing for instance the boundaries of GDP among all countries
in the same continent of the four selected countries, this
is more or less what we can do

```{r}
# give all the data to ggplot, we'll filter later
plt <- gapminder %>%
  ggplot() +
    geom_line(data = . %>%
      dplyr::filter(country %in% wanted_countries),
              aes(year, gdpPercap, color = country)) +
    # now group by continent and get the upper/lower bounds
    geom_ribbon(data = . %>%
      # min(NA) = NA, make sure NAs are excluded
      dplyr::filter(!is.na(gdpPercap)) %>%
      # gather all entries for each continent separately
      dplyr::group_by(continent, year) %>%
      # compute aggregated quantity (min/max)
      dplyr::summarize(minGdp = min(gdpPercap),
                       maxGdp = max(gdpPercap), across()) %>%
      dplyr::filter(country %in% wanted_countries),
              aes(ymin = minGdp, ymax = maxGdp,
                  x = year, color = country, fill = country),
              alpha = 0.1, linetype = "dashed", size = 0.2)
plt
```
But, since it looks a bit confusing, we might want four separate plots.

```{r}
wanted_continents <- gapminder %>%
  dplyr::filter(country %in% wanted_countries) %>%
  # extract one column from a dataframe (different from select)
  dplyr::pull(continent) %>%
  unique()

gapminder %>%
  dplyr::filter(continent %in% wanted_continents) %>%
  ggplot() +
    geom_line(data = . %>%
      dplyr::filter(country %in% wanted_countries),
              aes(year, gdpPercap, color = country)) +
    # now group by continent and get the upper/lower bounds
    geom_ribbon(data = . %>%
      # min(NA) = NA, make sure NAs are excluded
      dplyr::filter(!is.na(gdpPercap)) %>%
      # gather all entries for each continent separately
      dplyr::group_by(continent, year) %>%
      # compute aggregated quantity (min/max)
      dplyr::summarize(minGdp = min(gdpPercap),
                       maxGdp = max(gdpPercap), across()) %>%
      dplyr::filter(country %in% wanted_countries),
              aes(ymin = minGdp, ymax = maxGdp,
                  x = year, color = country, fill = country),
              alpha = 0.1, linetype = "dashed", size = 0.2) +
      facet_wrap(vars(continent)) +
      labs(title = paste("Country GDP per capita compared with",
           "continent lower and upper bounds"),
           x = "year", y = "GDP per capita (PPP dollars)")
```

## Probability

Base R provide functions to handle almost any probability distribution.
These functions are usually divided into four categories:

- density function
- distribution function
- quantile function
- random function (sampling)

```{r}
n <- 10
normal_samples <- rnorm(n = n, mean = 0, sd = 1) # sample 10 Gaussian samples
normal_samples

# compute the density function (over another Normal)
dnorm(normal_samples, mean = 2, sd = 1)
# cumulative distribution function
pnorm(normal_samples, mean = 0, sd = 1)

# get the quantiles of a normal
qnorm(c(0.05, 0.95), mean = 0, sd = 1)
```

## Extras

### File system and helper

R language provides several tools for management of files and function help.
Here some useful console commands. Note that most of them are also available
on RStudio through the graphic interface (buttons).

R saves all the variables and you can display them with `ls()`.
```{r}
rm(list = ls()) # clear up the space removing all variables stored so far
# let's add some variables
x <- 1:10
y <- x[x %% 2 == 0]

ls() # check variables in the environment
```

The working directory is the folder located on your computer from which R
navigates the filesystem.
```{r}
getwd() # check your wd
```
```{r, eval=FALSE}
setwd("./tmp") # set the working directory to an arbitrary (existing) folder
# save the current environment
save.image("./01_test.RData")
# check that it's on the working directory
dir()
```
RStudio typically save the environment automatically,
but sometimes (if not every time you close R) you should
clear the environment variables, because loading many variables 
when opening RStudio might fill up too much memory.

You can also read function helpers simply by typing `?function_name`. This
will open a formatted page with information about a specific R function or object.

```{r, eval=FALSE}
?quit # help for the quit function

?Arithmetic # help for more general syntax information
help(Trig) # or use help(name)
```

### Packages

Packages can be installed via command line using 
`install.packages("package_name")`, or through RStudio graphical interface.
```{r}
# the following function call is commented because package installation should
# not be included in a script (but you can find it commented, showing that the
# script requires a package as dependency)

# install.packages("tidyverse")

```
And then you can load the package with `library`.
```{r}
library(tidyverse)
```

### Arrow sign

The difference between `<-` and `=` is not just programming *style* preference.
Here an example where using `=` rather than `<-` makes a difference:

```{r, error=TRUE}
# gives error: argument 'b' is not part of 'within'
within(data.frame(a = rnorm(2)), b = a^2)
# 'b<-a^2' is the value passed to the expr argument of within()
within(data.frame(a = rnorm(2)), b <- a^2)
```
Although this event might never occur in one's programming experience, it's
safer (and more elegant) to use `<-` when assigning variable.

Besides, `->` is also valid and it is used (more intuitive) when assigning pipes
result to variables.

```{r,message=FALSE}
library(dplyr)

x <- starwars %>%
  dplyr::mutate(bmi = mass / ((height / 100) ^ 2)) %>%
  dplyr::filter(!is.na(bmi)) %>%
  dplyr::group_by(species) %>%
  dplyr::summarise(bmi = mean(bmi)) %>%
  dplyr::arrange(desc(bmi))

x
```

## Exercises

1. LogSumExp trick

Try to implement the log-sum-exp trick in a function that takes as argument
three `numeric` variables and computed the log of the sum of the exponentials
in a numerically stable way.
See this
[Wiki paragraph](https://en.wikipedia.org/wiki/LogSumExp#log-sum-exp_trick_for_log-domain_calculations)
if you don't know the trick yet.

```{r}
log_sum_exp3 <- function(a, b, c) {
  # delete this function and re-write it using the trick
  # in order to make it work
  return(log(exp(a) + exp(b) + exp(c)))
}

# test - this result is obviously wrong: edit the function above
log_sum_exp3(a = -1000, b = -1001, c = -999) # should give -998.5924
```

